Comparison to Christensen et al. 2018

### 2023-09-16 doc/paper-comparison.Rmd
no mark

---
title: "Comparison to Christensen et al. 2018"
author: "Renata Diaz and Hao Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{paper-comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  run_models: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=FALSE}
vers <- packageVersion("LDATS")
```

@@

This document provides a side-by-side comparison of **`LDATS`** (version `r vers`) results with analysis from [Christensen et al. 2018](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.2373).

@@

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("weecology/LDATS")
```

Load in the **`LDATS`** package.
```{r}
library(LDATS)
set.seed(42)
nseeds <- 200
nit <- 10000
```

@@

To change this functionality, you can re-render this file with:
```{r, eval = FALSE}
rmarkdown::render("paper-comparison.Rmd", params = list(run_models = TRUE))
```

@@

```{r set download location}
vignette_files <- tempdir()
```

To replicate the Christensen et al. 2018 analysis, we download some of the original scripts & data files from [Extreme-events-LDA repo](https://github.com/emchristensen/Extreme-events-LDA), as well as some raw data files from the [PortalData repo](https://github.com/weecology/PortalData):

@@

@@

```{r download scripts}
test_file <- file.path(vignette_files, "rodent_LDA_analysis.r")

if (!file.exists(test_file))
{
  # from the Extreme-events-LDA repo
  github_path <- "https://raw.githubusercontent.com/emchristensen/Extreme-events-LDA/master/"
  files_to_download <- c("rodent_LDA_analysis.r", "rodent_data_for_LDA.r", 
                         "AIC_model_selection.R", "changepointmodel.r", 
                         "LDA-distance.R", "Rodent_table_dat.csv", 
                         "LDA_figure_scripts.R")

  for (file in files_to_download)
  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, file))
  }

  # from the PortalData repo
  github_path <- "https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/"
  files_to_download <- c("moon_dates.csv", "Portal_rodent_trapping.csv")

  for (file in files_to_download)
  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, file))
  }
}
```

@@

Changepoint outputs

* `ldats_ldats.RDS`
  - the posterior distribution of changepoints, using the LDATS LDA model and the LDATS changepoint selection

* `ldats_paper.RDS`
  - the posterior distribution of changepoints, using the LDATS LDA model and the paper's changepoint selection

* `paper_ldats.RDS`
  - the posterior distribution of changepoints, using the paper LDA model and the LDATS changepoint selection

* `paper_paper.RDS`
  - the posterior distribution of changepoints, using the paper LDA model and the paper's changepoint selection

@@

```{r download pre-generated model outputs, eval = !params$run_model}
test_file <- file.path(vignette_files, "ldats_ldamodel.RDS")

if (!file.exists(test_file))
{
  # from the Extreme-events-LDA repo
  github_path <- "https://raw.githubusercontent.com/weecology/LDATS-replications/master/model_outputs/"
  files_to_download <- c("ldats_ldamodel.RDS", "paper_ldamodel.RDS", 
                         "ldats_ldats.RDS", "ldats_paper.RDS", 
                         "paper_ldats.RDS", "paper_paper.RDS", 
                         "lda_distances.png")

  for (file in files_to_download)
  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, file), 
                  mode = "wb")
  }
}
```

@@

```{r LDATS data}
data(rodents)

head(rodents[[1]])
```

@@

```{r Paper data}
# parameters for subsetting the full Portal rodents data
periods <- 1:436
control_plots <- c(2, 4, 8, 11, 12, 14, 17, 22)
species_list <- c("BA", "DM", "DO", "DS", "NA", "OL", "OT", "PB", "PE", "PF", 
                  "PH", "PI", "PL", "PM", "PP", "RF", "RM", "RO", "SF", "SH", "SO")

source(file.path(vignette_files, "rodent_data_for_LDA.r"))

@@

# assemble `paper_covariates`, the associated dates and covariate data
moondat <- read.csv(file.path(vignette_files, "moon_dates.csv"), stringsAsFactors = F)

@@

```{r rodent data comparison}
compare <- rodents[[1]] == paper_dat

length(which(rowSums(compare) < ncol(compare)))
```

There are 16 rows where the data included in LDATS differs from the paper data. This is because the LDATS data is not adjusted to account for trapping effort, while the paper data does, by dividing all census counts by the actual number of plots trapped and multiplying by 8 to account for incompletely-trapped censuses.

@@

```{}
  # retrieve data on number of plots trapped per month
  trap_table = read.csv('https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/Portal_rodent_trapping.csv')
  trap_table_controls = filter(trap_table, plot %in% selected_plots)
  nplots_controls = aggregate(trap_table_controls$sampled,by=list(period = trap_table_controls$period),FUN=sum)

@@

```{r adjust LDATS data after Christensen et al, eval = TRUE}
# get the trapping effort for each sample
trap_table <- read.csv(file.path(vignette_files, "Portal_rodent_trapping.csv"))
trap_table_controls <- dplyr::filter(trap_table, plot %in% control_plots)
nplots_controls <- aggregate(trap_table_controls$sampled, 
                            by = list(period = trap_table_controls$period), 
                            FUN = sum)

@@

Now we can compare the adjusted ldats dataset with both the original ldats dataset and the dataset from the paper:

```{r dataset comparisons}
compare_raw <- rodents[[1]] == ldats_rodents_adjusted
length(which(rowSums(compare_raw) < ncol(compare_raw)))

@@

```{r switch to adjusted rodents}
rodents[[1]] <- paper_dat
```

@@

```{r add dates to covariate table}
head(rodents$document_covariate_table)

new_cov_table <- dplyr::left_join(rodents$document_covariate_table, 
                                  dplyr::select(moondat, newmoonnumber, censusdate), 
                                  by = c("newmoon" = "newmoonnumber")) %>%
  dplyr::rename(date = censusdate)

@@

```{r LDATS LDAs, eval = params$run_models}
ldats_ldas <- LDATS::LDA_set(document_term_table = rodents$document_term_table, 
                             topics = 2:6, nseeds = nseeds)
ldats_ldamodel <- LDATS::select_LDA(LDA_models = ldats_ldas)[[1]]

@@

```{r paper LDAs, eval = params$run_models}
source(file.path(vignette_files, "AIC_model_selection.R"))
source(file.path(vignette_files, "LDA-distance.R"))

@@

png(file.path(vignette_files, "lda_distances.png"), width = 800, height = 400)
dat <- paper_dat # calculate_LDA_distance has some required named variables
best_seed <- calculate_LDA_distance(paper_dat, seeds_4topics)
dev.off()
mean_dist <- unlist(best_seed)[2]
max_dist <- unlist(best_seed)[3]

@@

```{r}
knitr::include_graphics(file.path(vignette_files, "lda_distances.png"))
```

@@

To visualize the LDA assignment of species to topics, we load in the saved LDA models from previously:
```{r}
ldamodel <- readRDS(file.path(vignette_files, "paper_ldamodel.RDS"))
ldats_ldamodel <- readRDS(file.path(vignette_files, "ldats_ldamodel.RDS"))
```

How the paper LDA model assigns species to topics:
```{r plot paper LDA, fig.width = 7, fig.height = 6}
plot(ldamodel, cols = NULL, option = "D")
```

How the LDATS LDA model assigns species to topics:
```{r plot LDATS LDA, fig.width = 7, fig.height = 6}
plot(ldats_ldamodel, cols = NULL, option = "D")
```

@@

```{r paper changepoint models}
#### Run changepoint ####
source(file.path(vignette_files, "changepointmodel.r"))

find_changepoints <- function(lda_model, paper_covariates, n_changepoints = 1:6)
{
  # set up parameters for model
  x <- dplyr::select(paper_covariates, 
                     year_continuous, 
                     sin_year, 
                     cos_year)

  # run models with 1, 2, 3, 4, 5, 6 changepoints
  cpt_results <- data.frame(n_changepoints = n_changepoints)
  cpt_results$cpt_model <- lapply(cpt_results$n_changepoints,
                                  function(n_changepoints)
                                  {
                                    changepoint_model(lda_model, x, n_changepoints, maxit = nit, 
                                                      weights = rep(1, NROW(x)))
                                  })
  return(cpt_results)
}

# Among a selection of models with different # of changepoints, 
#   - compute AIC
#   - select the model with the best AIC
#   - get the posterior distributions for the changepoints
select_cpt_model <- function(cpt_results, ntopics)
{
  # compute log likelihood as the mean deviance
  cpt_results$mean_deviances <- vapply(cpt_results$cpt_model, 
                                       function(cpt_model) {mean(cpt_model$saved_lls)}, 
                                       0)

@@

# transform the output from `compute_cpt` and match up the time indices with 
#   dates from the original data
get_dates <- function(cpt, covariates = paper_covariates)
{
  cpt$saved[,1,] %>%
    t() %>%
    as.data.frame() %>%
    reshape::melt() %>%
    dplyr::left_join(covariates, by = c("value" = "index"))
}
```

@@

Run the Christensen et al. time series model to identify changepoints on the LDA topics selected by LDATS:
```{r run LDATS LDA and paper cpt, eval = params$run_models}
ldats_paper_results <- find_changepoints(ldats_ldamodel, paper_covariates)

saveRDS(ldats_paper_results, file = file.path(vignette_files, "ldats_paper.RDS"))
```

Extract the dates of the changepoints:
```{r compute changepoints for LDATS LDA and paper cpt}
ldats_paper_results <- readRDS(file.path(vignette_files, "ldats_paper.RDS"))

@@

Run the Christensen et al. time series model to identify changepoints on the LDA topics selected by Christensen et al.:
```{r run paper LDA and paper cpt, eval = params$run_models}
paper_paper_results <- find_changepoints(ldamodel, paper_covariates)

@@

Extract the dates of the changepoints:
```{r compute changepoints for paper LDA and paper cpt}
paper_paper_results <- readRDS(file.path(vignette_files, "paper_paper.RDS"))

@@

```{r run LDATS LDA and LDATS cpt, eval = params$run_models}
ldats_ldats_results <- TS_on_LDA(LDA_models = ldats_ldamodel,
                                 document_covariate_table = rodents$document_covariate_table,
                                 formulas = ~ sin_year + cos_year,
                                 nchangepoints = 1:6,
                                 timename = "newmoon",
                                 weights = NULL,
                                 control = list(nit = nit))

saveRDS(ldats_ldats_results, file = file.path(vignette_files, "ldats_ldats.RDS"))
```

@@

```{r construct lookup table for LDATS output for changepoint times}
# make the full sequence of possible newmoon values
full_index <- seq(min(rodents$document_covariate_table$newmoon), 
                  max(rodents$document_covariate_table$newmoon))

@@

```{r compute changepoints for LDATS LDA and LDATS cpt}
ldats_ldats_results <- readRDS(file.path(vignette_files, "ldats_ldats.RDS"))

@@

```{r run paper LDA and LDATS cpt, eval = params$run_models}
paper_ldats_results <- TS_on_LDA(LDA_models = ldamodel,
                             document_covariate_table = rodents$document_covariate_table,
                             formulas = ~ sin_year + cos_year,
                             nchangepoints = 1:6,

@@

saveRDS(paper_ldats_results, file = file.path(vignette_files, "paper_ldats.RDS"))
```

@@

```{r select paper lda + ldats cpt}
paper_ldats_results <- readRDS(file.path(vignette_files, "paper_ldats.RDS"))

@@

```{r}
nlevels(ldats_paper_cpt_dates$variable)
nlevels(paper_paper_cpt_dates$variable)
nlevels(ldats_ldats_cpt_dates$variable)
nlevels(paper_ldats_cpt_dates$variable)
```

@@

```{r plot paper LDA and LDATS cpts, fig.width = 7, fig.height = 6}
plot(paper_ldats_cpt)
```

### LDATS LDA and LDATS changepoint
```{r plot ldats LDA and LDATS cpt, fig.width = 7, fig.height = 6}
plot(ldats_ldats_cpt)
```

### Paper LDA and paper changepoint
```{r plot paper LDA and paper cpt, fig.width = 7, fig.height = 6}
paper_cpts <- find_changepoint_location(paper_paper_cpt)
ntopics <- ldamodel@k

paper_cpt_plot <- get_ll_non_memoized_plot(ldamodel, paper_covariates, paper_cpts, make_plot = TRUE,
                                           weights = rep(1, NROW(paper_covariates)))

annual_hist(paper_paper_cpt, paper_covariates$year_continuous)
paper_cpt_plot
```

### LDATS LDA and paper changepoint
```{r plot LDATS lda and paper cpt, fig.width = 7, fig.height = 6}
ldats_cpts <- find_changepoint_location(ldats_paper_cpt)
ntopics <- ldats_ldamodel@k

ldats_cpt_plot <- get_ll_non_memoized_plot(ldats_ldamodel, paper_covariates, ldats_cpts, make_plot = TRUE,
                                           weights = rep(1, NROW(paper_covariates)))

annual_hist(ldats_paper_cpt, paper_covariates$year_continuous)
ldats_cpt_plot
```

@@

## Report changepoint dates
```{r report cpt dates, include = F}
cpt_dates <- dplyr::bind_rows("paperLDA_paperCPT" = paper_paper_cpt_dates, 
                              "ldatsLDA_paperCPT" = ldats_paper_cpt_dates, 
                              "ldatsLDA_ldatsCPT" = ldats_ldats_cpt_dates, 
                              "paperLDA_ldatsCPT" = paper_ldats_cpt_dates, 
                              .id = "analysis") %>%
  dplyr::group_by(analysis, variable) %>%
  dplyr::summarize(date = mean(date)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(changepoint = variable) %>%
  tidyr::spread(analysis, date)
```

```{r print cpt dates}
knitr::kable(cpt_dates)
```

>>>>

---
title: "Comparison to Christensen et al. 2018"
author: "Renata Diaz, Juniper Simonis, and Hao Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{paper-comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  run_models: FALSE
---

@@

This document provides a side-by-side comparison of **`LDATS`** (version 0.3.0) results with analysis from [Christensen et al. 2018](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.2373). 
Due to the size and duration of model runs, we use pre-generated model output from the [LDATS-replications repo](https://github.com/weecology/LDATS-replications).

@@

```r
install.packages("devtools")
devtools::install_github("weecology/LDATS")
```

Load in the **`LDATS`** package.

```r
library(LDATS)
set.seed(42)
nseeds <- 200
nit <- 10000
```

To run the analyses here, you will also need to download **`dplyr`**, **`gridExtra`**, **`multipanel`**, **`RColorBrewer`**, **`RCurl`**, and **`reshape2`** as the manuscript's code relies on these packages.

```r
install.packages(c("dplyr", "gridExtra", "multipanel", "RColorBrewer", "RCurl", "reshape2"))
```

@@

To change this functionality, you can re-render this file with:

```r
rmarkdown::render("paper-comparison.Rmd", params = list(run_models = TRUE))
```

@@

```r
vignette_files <- "."
```

To replicate the Christensen et al. 2018 analysis, we download some of the original scripts & data files from [Extreme-events-LDA repo](https://github.com/emchristensen/Extreme-events-LDA), as well as some raw data files from the [PortalData repo](https://github.com/weecology/PortalData), which are stored in the [LDATS-replications repo](https://github.com/weecology/LDATS-replications):

@@

* `paper_dat.csv`
  - rodent data table from Christensen et al. 2018

* `paper_dates.csv`
  - dates used in Christensen et al. 2018

* `paper_covariates.csv`
  - table of dates and covariate data

@@

```r
test_file <- file.path(vignette_files, "scripts", "rodent_LDA_analysis.r")

if (!file.exists(test_file)){

  # scripts
  dir.create(file.path(vignette_files, "scripts"), showWarnings = FALSE)
  github_path <- "https://raw.githubusercontent.com/weecology/LDATS-replications/master/scripts/"
  files_to_download <- c("rodent_LDA_analysis.r", "rodent_data_for_LDA.r", 
                         "AIC_model_selection.R", "changepointmodel.r", 
                         "LDA-distance.R", "LDA_figure_scripts.R")

  for (file in files_to_download)  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, "scripts", file))
  }

  # data
  dir.create(file.path(vignette_files, "data"), showWarnings = FALSE)
  github_path <- "https://raw.githubusercontent.com/weecology/LDATS-replications/master/data/"
  files_to_download <- c("moon_dates.csv", "Portal_rodent_trapping.csv", 
                         "Rodent_table_dat.csv", "paper_dat.csv",
                         "paper_dates.csv", "paper_covariates.csv")

  for (file in files_to_download)  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, "data", file))
  }
}
```

@@

Changepoint outputs:

* `ldats_ldats.RDS`, `ldats_ldats_cpt.RDS`, `ldats_ldats_cpt_dates.RDS`
  - the posterior distribution, count, and dates of changepoints, using the LDATS LDA model and the LDATS changepoint selection

* `ldats_paper.RDS`, `ldats_paper_cpt.RDS`, `ldats_paper_cpt_dates.RDS
  - the posterior distribution, count, and dates of changepoints, using the LDATS LDA model and the paper's changepoint selection

* `paper_ldats.RDS`, `paper_ldats_cpt.RDS`, `paper_ldats_cpt_dates.RDS
  - the posterior distribution, count, and dates of changepoints, using the paper LDA model and the LDATS changepoint selection

* `paper_paper.RDS`, `paper_paper_cpt.RDS`, `paper_paper_cpt_dates.RDS
  - the posterior distribution, count, and dates of changepoints, using the paper LDA model and the paper's changepoint selection

* `cpt_dates.RDS`
  - summary table of changepoint results across models

@@

* `paper_paper_cpt_plot.png`
  - figure showing the time series results for the paper analysis of the paper LDA

* `ldats_paper_cpt_plot.png`
  - figure showing the time series results for the paper analysis of the LDATS LDA

* `annual_hist.RDS`
  - function for making histogram of change points over years

```r
test_file <- file.path(vignette_files, "output", "ldats_ldamodel.RDS")

if (!file.exists(test_file)){

  dir.create(file.path(vignette_files, "output"), showWarnings = FALSE)
  github_path <- "https://raw.githubusercontent.com/weecology/LDATS-replications/master/output/"
  files_to_download <- c("ldats_ldamodel.RDS", "paper_ldamodel.RDS", 
                         "ldats_ldats.RDS", "ldats_paper.RDS", 
                         "paper_ldats.RDS", "paper_paper.RDS", 
                         "ldats_rodents_adjusted.RDS", "rodents.RDS",
                         "ldats_paper_cpt.RDS", "ldats_paper_cpt_dates.RDS",
                         "ldats_ldats_cpt.RDS", "ldats_ldats_cpt_dates.RDS",
                         "paper_paper_cpt.RDS", "paper_paper_cpt_dates.RDS",
                         "paper_ldats_cpt.RDS", "paper_ldats_cpt_dates.RDS",
                         "annual_hist.RDS", "cpt_dates.RDS",
                         "lda_distances.png", "paper_paper_cpt_plot.png",
                         "ldats_paper_cpt_plot.png")

  for (file in files_to_download){
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, "output", file), 
                  mode = "wb")
  }
}
```

@@

```r
data(rodents)

head(rodents[[1]])
#>   BA DM DO DS NA. OL OT PB PE PF PH PI PL PM PP RF RM RO SF SH SO
#> 1  0 13  0  2   2  0  0  0  1  1  0  0  0  0  3  0  0  0  0  0  0
#> 2  0 20  1  3   2  0  0  0  0  4  0  0  0  0  2  0  0  0  0  0  0
#> 3  0 21  0  8   4  0  0  0  1  2  0  0  0  0  1  0  0  0  0  0  0
#> 4  0 21  3 12   4  2  3  0  1  1  0  0  0  0  0  0  0  0  0  0  0
#> 5  0 16  1  9   5  2  1  0  0  2  0  0  0  0  0  0  1  0  0  0  0
#> 6  0 17  1 13   5  1  5  0  0  3  0  0  0  0  0  0  0  0  0  0  0
```

@@

```r
# parameters for subsetting the full Portal rodents data
periods <- 1:436
control_plots <- c(2, 4, 8, 11, 12, 14, 17, 22)
species_list <- c("BA", "DM", "DO", "DS", "NA", "OL", "OT", "PB", "PE", "PF", 
                  "PH", "PI", "PL", "PM", "PP", "RF", "RM", "RO", "SF", "SH", "SO")

source(file.path(vignette_files, "scripts", "rodent_data_for_LDA.r"))

@@

# assemble `paper_covariates`, the associated dates and covariate data
moondat <- read.csv(file.path(vignette_files, "data", "moon_dates.csv"), stringsAsFactors = F)

@@

```r
compare <- rodents[[1]] == paper_dat

length(which(rowSums(compare) < ncol(compare)))
#> [1] 16
```

There are 16 rows where the data included in LDATS differs from the paper data. This is because the LDATS data is not adjusted to account for trapping effort, while the paper data is, by dividing all census counts by the actual number of plots trapped and multiplying by 8 to account for incompletely-trapped censuses.

@@

```r
  # retrieve data on number of plots trapped per month
  trap_table = read.csv('https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/Portal_rodent_trapping.csv')
  trap_table_controls = filter(trap_table, plot %in% selected_plots)
  nplots_controls = aggregate(trap_table_controls$sampled,by=list(period = trap_table_controls$period),FUN=sum)

@@

```r
# get the trapping effort for each sample
trap_table <- read.csv(file.path(vignette_files, "data", "Portal_rodent_trapping.csv"))
trap_table_controls <- dplyr::filter(trap_table, plot %in% control_plots)
nplots_controls <- aggregate(trap_table_controls$sampled, 
                            by = list(period = trap_table_controls$period), 
                            FUN = sum)

@@

Now we can compare the adjusted LDATS dataset with both the original ldats dataset and the dataset from the paper:

```r
compare_raw <- rodents[[1]] == ldats_rodents_adjusted
length(which(rowSums(compare_raw) < ncol(compare_raw)))

@@

```r
rodents[[1]] <- paper_dat
```

@@

```r
head(rodents$document_covariate_table)
#>   newmoon sin_year cos_year
#> 1       1  -0.2470  -0.9690
#> 2       2  -0.6808  -0.7325
#> 3       3  -0.9537  -0.3008
#> 4       4  -0.9813   0.1925
#> 5       5  -0.7583   0.6519
#> 6       6  -0.3537   0.9354
```

```r
new_cov_table <- dplyr::left_join(rodents$document_covariate_table, 
                                  dplyr::select(moondat, newmoonnumber, censusdate), 
                                  by = c("newmoon" = "newmoonnumber")) %>%
                                  dplyr::rename(date = censusdate)

@@

```r
ldats_ldas <- LDATS::LDA_set(document_term_table = rodents$document_term_table, 
                             topics = 2:6, nseeds = nseeds)
ldats_ldamodel <- LDATS::select_LDA(LDA_models = ldats_ldas)[[1]]

@@

```r
source(file.path(vignette_files, "scripts", "AIC_model_selection.R"))
source(file.path(vignette_files, "scripts", "LDA-distance.R"))

@@

png(file.path(vignette_files, "output", "lda_distances.png"), width = 800, height = 400)
dat <- paper_dat # calculate_LDA_distance has some required named variables
best_seed <- calculate_LDA_distance(paper_dat, seeds_4topics)
dev.off()
mean_dist <- unlist(best_seed)[2]
max_dist <- unlist(best_seed)[3]

@@

```r
knitr::include_graphics(file.path(vignette_files, "output", "lda_distances.png"))
```

![plot of chunk unnamed-chunk-10](./output/lda_distances.png)

@@

To visualize the LDA assignment of species to topics, we load in the saved LDA models from previously:

```r
ldamodel <- readRDS(file.path(vignette_files, "output", "paper_ldamodel.RDS"))
ldats_ldamodel <- readRDS(file.path(vignette_files, "output", "ldats_ldamodel.RDS"))
```

How the paper LDA model assigns species to topics:

```r
plot(ldamodel, cols = NULL, option = "D")
```

![plot of chunk plot-paper-LDA](figure/plot-paper-LDA-1.png)

How the LDATS LDA model assigns species to topics:

```r
plot(ldats_ldamodel, cols = NULL, option = "D")
```

![plot of chunk plot-LDATS-LDA](figure/plot-LDATS-LDA-1.png)

@@

```r
#### Run changepoint ####
source(file.path(vignette_files, "scripts", "changepointmodel.r"))

find_changepoints <- function(lda_model, paper_covariates, n_changepoints = 1:6){
  # set up parameters for model
  x <- dplyr::select(paper_covariates, 
                     year_continuous, 
                     sin_year, 
                     cos_year)

  # run models with 1, 2, 3, 4, 5, 6 changepoints
  cpt_results <- data.frame(n_changepoints = n_changepoints)
  cpt_results$cpt_model <- lapply(cpt_results$n_changepoints,
                                  function(n_changepoints){
                                    changepoint_model(lda_model, x, n_changepoints, maxit = nit, 
                                                      weights = rep(1, NROW(x)))
                                  })
  return(cpt_results)
}

# Among a selection of models with different # of changepoints, 
#   - compute AIC
#   - select the model with the best AIC
#   - get the posterior distributions for the changepoints
select_cpt_model <- function(cpt_results, ntopics){
  # compute log likelihood as the mean deviance
  cpt_results$mean_deviances <- vapply(cpt_results$cpt_model, 
                                       function(cpt_model) {mean(cpt_model$saved_lls)}, 
                                       0)

@@

# transform the output from `compute_cpt` and match up the time indices with 
#   dates from the original data
get_dates <- function(cpt, covariates = paper_covariates){
  cpt$saved[,1,] %>%
    t() %>%
    as.data.frame() %>%
    reshape::melt() %>%
    dplyr::left_join(covariates, by = c("value" = "index"))
}
```

@@

Run the Christensen et al. time series model to identify changepoints on the LDA topics selected by LDATS:

```r
ldats_paper_results <- find_changepoints(ldats_ldamodel, paper_covariates)

saveRDS(ldats_paper_results, file = file.path(vignette_files, "output", "ldats_paper.RDS"))
```

Extract the dates of the changepoints:

```r
ldats_paper_results <- readRDS(file.path(vignette_files, "output", "ldats_paper.RDS"))

@@

Run the Christensen et al. time series model to identify changepoints on the LDA topics selected by Christensen et al.:

```r
paper_paper_results <- find_changepoints(ldamodel, paper_covariates)

@@

Extract the dates of the changepoints:

```r
paper_paper_results <- readRDS(file.path(vignette_files, "output", "paper_paper.RDS"))

@@

```r
ldats_ldats_results <- TS_on_LDA(LDA_models = ldats_ldamodel,
                                 document_covariate_table = rodents$document_covariate_table,
                                 formulas = ~ sin_year + cos_year,
                                 nchangepoints = 1:6,
                                 timename = "newmoon",
                                 weights = NULL,
                                 control = list(nit = nit))

saveRDS(ldats_ldats_results, file = file.path(vignette_files, "output", "ldats_ldats.RDS"))
```

@@

```r
# make the full sequence of possible newmoon values
full_index <- seq(min(rodents$document_covariate_table$newmoon), 
                  max(rodents$document_covariate_table$newmoon))

@@

```r
ldats_ldats_results <- readRDS(file.path(vignette_files, "output", "ldats_ldats.RDS"))

@@

```r
paper_ldats_results <- TS_on_LDA(LDA_models = ldamodel,
                             document_covariate_table = rodents$document_covariate_table,
                             formulas = ~ sin_year + cos_year,
                             nchangepoints = 1:6,

@@

saveRDS(paper_ldats_results, file = file.path(vignette_files, "output", "paper_ldats.RDS"))
```

@@

```r
paper_ldats_results <- readRDS(file.path(vignette_files, "output", "paper_ldats.RDS"))

@@

```r
nlevels(ldats_paper_cpt_dates$variable)
#> [1] 4
nlevels(paper_paper_cpt_dates$variable)
#> [1] 4
nlevels(ldats_ldats_cpt_dates$variable)
#> [1] 4
nlevels(paper_ldats_cpt_dates$variable)
#> [1] 4
```

@@

```r
plot(paper_ldats_cpt)
```

![plot of chunk plot paper LDA and LDATS cpts](figure/plot paper LDA and LDATS cpts-1.png)

```r
plot(ldats_ldats_cpt)
```

![plot of chunk plot ldats LDA and LDATS cpt](figure/plot ldats LDA and LDATS cpt-1.png)

```r
annual_hist <- readRDS(file.path(vignette_files, "output", "annual_hist.RDS"))
```

```r
paper_cpts <- find_changepoint_location(paper_paper_cpt)
ntopics <- ldamodel@k

png(file.path(vignette_files, "output", "paper_paper_cpt_plot.png"), width = 800, height = 600)
get_ll_non_memoized_plot(ldamodel, paper_covariates, paper_cpts, make_plot = TRUE,
                                           weights = rep(1, NROW(paper_covariates)))
dev.off()
```

```r
paper_paper_hist <- annual_hist(paper_paper_cpt, paper_covariates$year_continuous)
```

![plot of chunk plot paper LDA and LDATS cpts2](figure/plot paper LDA and LDATS cpts2-1.png)

```r
knitr::include_graphics(file.path(vignette_files, "output", "paper_paper_cpt_plot.png"))
```

![plot of chunk unnamed-chunk-19](./output/paper_paper_cpt_plot.png)

```r
ldats_cpts <- find_changepoint_location(ldats_paper_cpt)
ntopics <- ldats_ldamodel@k

png(file.path(vignette_files, "output", "ldats_paper_cpt_plot.png"), width = 800, height = 600)
get_ll_non_memoized_plot(ldats_ldamodel, paper_covariates, ldats_cpts, make_plot = TRUE,
                                           weights = rep(1, NROW(paper_covariates)))
dev.off()
```

```r
ldats_paper_hist <- annual_hist(ldats_paper_cpt, paper_covariates$year_continuous)
```

![plot of chunk plot LDATS lda and paper cpt2](figure/plot LDATS lda and paper cpt2-1.png)

```r
knitr::include_graphics(file.path(vignette_files, "output", "ldats_paper_cpt_plot.png"))
```

![plot of chunk unnamed-chunk-20](./output/ldats_paper_cpt_plot.png)

@@

## Report changepoint dates

```r
knitr::kable(cpt_dates)
```

|changepoint |ldatsLDA_ldatsCPT |ldatsLDA_paperCPT |paperLDA_ldatsCPT |paperLDA_paperCPT |
|:-----------|:-----------------|:-----------------|:-----------------|:-----------------|
|V1          |1984-01-30        |1984-01-11        |1984-04-27        |1984-01-11        |
|V2          |1991-05-30        |1991-05-29        |1993-04-19        |1991-05-29        |
|V3          |1999-01-25        |1998-12-21        |1999-07-07        |1998-12-21        |
|V4          |2009-11-15        |2009-09-29        |2010-01-20        |2009-09-29        |
